\chapter*{Abstract}\thispagestyle{rmplain}
\addcontentsline{toc}{chapter}{Abstract}
\label{05_abstract}

This work focuses on a proof of concept approach to evaluate and improve building footprints in 5 study areas () in the global south with the base dataset that should be improved google open buildings. In those area building footprints are important for mapping, humanitarian work, planning and geoospatial anlysis. In a lot of regions i nthe global south there is still no contain geometric errors the building footprints but are already used in certain workflows. Therefore the approach of this work is not to create a new model to generate a new dataset but aims to improve a widly free usable dataset for this reagion and improve that locally. The google dataset has various challenges still that are also listed in there technical report such as alignmend, completnes and gemetrical accurancy in difficult settings.  For this approach the emerging technologies of Foundation models are used. The methodology difference is that a two step framework is used with an Multimodal large language model and Segment Anything 3 from sam. So evaluation and correction can be divided where the mllm is responsiblbe for quality assessment and routing of the workflow and sam is responsibke for the correction of the polygons that wehere detected as error invluenced. Afterward the generated geometries needed a postprocessing becuase sam outputs raster pixel based segmentations masks so the pypi package building_reguliser got adjusted to work with my sam poylgons in global south geprägt from incremently grown buildings with many different corners edges etc. The reults got evaaluated 1256 polygons manually classified bevoir and after the workflow. Here the workflow corrected best simpler geometry buildings in mozambique and niger with xx and xx, in liberia xx and in mexico and nepal with most complex building geometries xx and xx. And in total xx. What didn worked was it was used a qwen 3 8billion parameter model hosted on a runpod gpu with an a40 this mllm was able to decide if errors are present and if all the house is in the picture but it struggled with fine grained comparison tasks such as comparing befor and after if it got better and also to generate semantic error descriptions between the high qulity imagery and the google open building polygon. it also generated point placment coordinates inside and outside the poylgon. 
The interpretation out of this work is that foundation models