\chapter{Discussion}
\label{09_diskussion}


\section{Scope and Overall Effectiveness of the Refinement Workflow}

The results of this thesis should be understood as an evaluation of the update-oriented workflow introduced in Section~\ref{Introduction} \citep{Qin_changedetection, Wadembere_GeometryAdjustment}, rather than as a conventional building-footprint extraction benchmark. The aim was not to create a new building dataset from scratch or to replace existing global building-footprint products. Instead, the objective was to test whether already available polygons can be visually assessed and selectively improved using recent vision and multimodal AI methods.

This perspective distinguishes the proposed workflow from most building-footprint extraction approaches. Whereas many existing methods derive building footprints directly from high-resolution imagery \citep{Luo_overviewbuioldingfootprint,rastogi2022automatic}, this work starts from an existing polygon dataset. The workflow therefore focuses on local correction: a multimodal large language model is used to support image-based quality assessment and workflow decisions, while SAM is used to generate refined geometry candidates for polygons that show visible disagreement with the target imagery.

This is particularly relevant for heterogeneous and partly informal settlement structures in the Global South, where available building datasets can differ in completeness, geometric consistency, and local suitability \citep{sirko2021continentalscalebuildingdetectionhigh,CHAMBERLAIN2024102104}. The discussion therefore focuses on where the proof-of-concept workflow produced meaningful improvements, where the MLLM, SAM, or post-processing introduced new errors, and what this implies for future human-in-the-loop geospatial data maintenance workflows.

A central limitation has to be kept in mind throughout this interpretation. The OpenAerialMap imagery used for evaluation was not necessarily identical to the imagery used to generate the original Google Open Buildings polygons. Consequently, temporal differences, image registration effects, orthorectification uncertainty, and georeferencing errors may contribute to apparent polygon-image discrepancies \citep{Toutin01052004}. The results should therefore be interpreted as improvements relative to the selected evaluation imagery, not as proof of absolute geodetic correctness. This limitation is discussed in more detail in Section~\ref{sec:limitations_proof_of_concept}.

\subsection{Overall improvement of the evaluated polygons}
\label{sec:discussion_overall_effectiveness}

The manually evaluated results indicate that the workflow improved the visual agreement between the input polygons and the selected OpenAerialMap imagery in all five study areas. The benchmark compared polygons from the Google Open Buildings dataset with high-resolution OpenAerialMap imagery \citep{sirko2021continentalscalebuildingdetectionhigh,googleopenbuildings,openaerialmap}. Across the manually evaluated subset, 82.8\% of the polygons received a better quality rating after post-processing, while 13.1\% remained largely unchanged and 4.1\% received a lower rating (Figure~\ref{fig:country_improvement_rates}). The share of polygons rated \textit{good} or \textit{perfect} increased from a very low initial level to 77.8\% after post-processing (Table~\ref{tab:country_summary}).

This improvement has to be interpreted in relation to the initially low quality of the input data. In the original polygons, only 1.6\% were rated above \textit{ok}, while 98.4\% were rated as \textit{ok} or \textit{bad} (Figure~\ref{fig:transition_heatmap}). In addition, 1,283 polygons were removed from the starting dataset during MLQA because no corresponding visible building could be identified reliably in the selected imagery (Table~\ref{tab:mlqa_overview}). The workflow was therefore most effective in a setting where many input polygons showed visible disagreement with the target imagery. This is an important distinction: the method should not be interpreted as a universal improvement tool for all building datasets, but as a selective correction workflow for polygons that are likely to contain visible geometric or positional errors.

\section{Corrected Errors and Remaining Challenges}

This section examines how the composition of geometric errors changed during the improvement workflow. According to the manually assigned error labels, 77.3\% of the original Google Open Buildings polygons were classified as \texttt{SHIFTED} when compared with the corresponding building roofs visible in the OpenAerialMap TIFF imagery. After post-processing, this error category was reduced to 4.3\%, making positional displacement the error type that was corrected most successfully by the workflow (Figure~\ref{fig:error_profile_by_location}).

\subsection{Positional shifts and remote-sensing uncertainty}

The \texttt{SHIFTED} category should not be interpreted as evidence of a single systematic displacement affecting all polygons equally. As shown in the spatial shift analysis in Figure~\ref{fig:error_profile_shift}, the observed shifts did not follow a consistent direction or distance. Liberia, Mexico, and Nepal showed broadly distributed shift vectors, while Niger and Mozambique were more concentrated but still not homogeneous. Nepal had the largest displacement spread, with a 90th percentile of 8.88\,m, followed by Mexico at 7.20\,m and Liberia at 4.68\,m. A simple global translation of the complete dataset would therefore not have been sufficient to align all polygons with the imagery.

As introduced in Section~\ref{sec:geospatial_data_quality}, differences between vector geometries and visible building locations can be introduced at several stages of the remote-sensing acquisition and processing chain \citep{Toutin01052004}. These effects are particularly relevant for high-resolution and off-nadir imagery, where even relatively small geometric inconsistencies can become clearly visible at the scale of individual buildings.

In principle, many of these distortions should be reduced during orthorectification. However, the necessary information on terrain models, ground control points, sensor models, and image-processing parameters was not available for the imagery used to generate the Google Open Buildings polygons or for every OpenAerialMap image used in this thesis. It was therefore not possible to determine whether the observed displacement originated from the Open Buildings source imagery, the OpenAerialMap orthomosaics, differences between the two image products, temporal change in the built environment, or a combination of these factors.

For this reason, the workflow does not attempt to correct the remote-sensing imagery itself. Instead, the individual OpenAerialMap TIFFs are treated as the visual reference to which the vector geometries are adapted. The workflow therefore improves the internal alignment between building polygons and the selected imagery, but it does not prove that the resulting geometries are more accurate in an absolute geodetic sense. In a practical GIS workflow, however, this image-specific adaptation can still be useful when the target imagery is known and represents the dataset against which the building polygons will be used or evaluated.

Although shifted polygons were corrected successfully, they also introduced one of the largest technical challenges during workflow development. When an original polygon was strongly displaced, a patch cropped directly around its geometry did not always contain the complete visible building. This was one of the main reasons for separating the processing workflow into \textit{partial-house} and \textit{full-house} cases. The partial-house workflow allowed the model to search beyond the immediate extent of the original polygon and reconstruct building sections that would otherwise have been excluded from the image patch.

\section{Transferability Across Study Areas and Reference Data}

\textbf{Niger and Mozambique:}

Clear performance differences were observed between the study areas. Niger and Mozambique achieved the strongest overall results, with 85.9\% and 83.8\% of buildings reaching the \textit{good} or \textit{perfect} categories after post-processing (Table~\ref{tab:country_summary}). Both study areas had the advantage that many buildings showed comparatively simple and compact shapes. However, the contrast analysis shows that their visual image conditions were not the same. Niger had the strongest building-to-surrounding contrast of all study areas, with a median contrast value of 2.053. Mozambique, in contrast, had by far the lowest measured contrast, with a median value of only 0.246 (Table~\ref{tab:building_surrounding_contrast}).

The high final performance in Mozambique suggests that simple building morphology can partly compensate for low roof-ground contrast once a building has entered the refinement workflow. At the same time, Mozambique also had the highest number of buildings classified as \textit{No visible} during MLQA, with 424 removed cases. This indicates that low contrast may have affected the initial visibility decision more strongly than the later geometric refinement of successfully routed buildings.

The Qwen model was responsible for deciding whether a visible building was present. This decision may depend not only on image quality, but also on whether similar building types were sufficiently represented during model training. Since the buildings in Mozambique were among the most rudimentary structures in the dataset, it is possible that the model was affected by the type of geographic representation bias described in the theoretical background \citep{shankar2017classificationrepresentationassessinggeodiversity,yin2023givlimprovinggeographicalinclusivity}. This suggests that the decision performance of the MLLM may not only depend on contrast or resolution, but also on whether the visual appearance of buildings in a specific region is well represented in the model's training data.

\textbf{Liberia:}

Liberia represents a somewhat separate case within this work. The study area had favourable characteristics for object detection, although the starting polygons were still dominated by low and medium manual quality ratings. Liberia showed moderate visual contrast, with a median contrast value of 0.838, and the highest texture difference between buildings and surroundings, with a median texture-difference value of 28.0 (Table~\ref{tab:building_surrounding_contrast}). Many buildings also appeared as individually standing structures rather than as densely attached buildings.

In addition, Liberia was the study area most frequently used during the workflow development and visual inspection. Several small workflow adjustments, additional features, and prompt refinements were developed based on intermediate results from the Liberia study area. This may have slightly favoured the results for Liberia. At the same time, Liberia was geometrically more demanding than the two best-performing study areas: the average vertex count increased by 154.6\% after post-processing, compared with 45.0\% in Niger and 43.1\% in Mozambique (Table~\ref{tab:country_summary}). The Liberia results should therefore be interpreted as partly influenced by favourable image and settlement characteristics, but also by the fact that the workflow was developed and visually inspected more often in this area.

\textbf{Mexico and Nepal:}

Mexico and Nepal performed worse compared with the other study areas. Their final \textit{good} or \textit{perfect} shares were 72.4\% and 65.6\%, respectively, and both areas showed substantially larger geometry changes than Niger and Mozambique. Mexico had a median absolute area change of 30.7\%, while Nepal reached 33.4\%. Nepal also showed the largest median positional correction at 5.1\,m and the strongest increase in average vertex count at 382.5\% (Table~\ref{tab:country_summary}). These results suggest that the workflow had to substantially reconstruct many polygons rather than only adjust their position.

The lower performance in Mexico and Nepal was not only a question of image quality, but also of structural complexity. During visual interpretation, it became clear that many buildings in Nepal had roof terraces with small additional structures on top, often built with different roofing materials. In these cases, the model frequently appeared to struggle with the decision of whether it should segment the complete building including the roof terrace, or only the smaller rooftop structure. A similar pattern was observed in Mexico, where buildings often appeared as irregular combinations of extensions, added rooms, and mixed materials.

This type of building morphology can be described as incremental housing. In such housing forms, buildings are not necessarily constructed as one complete and clearly bounded object, but are extended, modified, and upgraded over time by their residents. This can result in irregular outlines, roof terraces, small rooftop additions, and heterogeneous roofing materials. Especially in Mexico and Nepal, this incremental and heterogeneous building style likely increased the difficulty of the segmentation task. The model had to decide where the visible building footprint begins and ends, and whether attached or stacked structures should be treated as part of the same building or as separate objects. This may explain why these two areas performed worse, even though their challenges were not only related to image quality, but also to ambiguous building morphology and structural complexity.

Overall, the central result is that the workflow improved most manually evaluated polygons, but not equally in every study area. The quality transition and improvement-rate results therefore have to be interpreted in relation to initial polygon quality, image quality, local contrast, settlement morphology, vegetation, and the structural complexity of the buildings.

\begin{itemize}
    \item Niger and Mozambique had the strongest final results and more stable area preservation.
    \item Liberia and Mexico were intermediate but differed in vegetation context, texture contrast, and area changes.
    \item Nepal had the weakest final quality, strongest area changes, and highest geometry complexity increase.
    \item Transferability is possible, but strongly depends on settlement morphology, image quality, vegetation, and initial polygon quality.
\end{itemize}

\subsection{Comparison with the Switzerland reference test}

As a small additional analysis, the same workflow that was developed for the Global South study areas and the Google Open Buildings polygons was also applied to Microsoft building polygons in Switzerland. This test was intended as a contrast case, because the Microsoft buildings already showed high agreement with the official TLM dataset, reaching an F1-score of 0.766 at the strictest IoU threshold of 0.50 (Table~\ref{tab:switzerland_reference_evaluation}). In this setting, the workflow was not able to improve the polygons according to the selected metrics. Instead, the regularized output performed slightly worse than the Microsoft source polygons when compared with the TLM reference, especially at the strictest IoU threshold.

A likely reason is that the workflow was mainly designed for cases where the initial polygons contain visible geometric errors and can therefore benefit from image-based refinement. In the Swiss test area, however, the input data were already of comparatively high quality. As a result, small changes introduced by SAM and the subsequent regularization were more likely to reduce the agreement with the reference data than to improve it. In some cases, neighbouring buildings were merged although they were separate in the reference dataset. In other cases, buildings that belonged together were split into separate parts. These changes reduced the precision of the output.

The comparison is further complicated by the fact that the three datasets were not necessarily derived from the same imagery or the same geometric definition of a building. The TLM dataset does not always follow the visible roof outline exactly, because it represents authoritative building footprints rather than image-derived roof polygons. In contrast, the image-based refinement workflow tends to adapt polygons towards the visible roof structure in the selected imagery. Therefore, a lower IoU against TLM does not necessarily mean that the resulting polygon is visually worse with respect to the image, but it does show that the workflow is not automatically suitable for improving authoritative footprint data. Overall, the Switzerland reference test supports the interpretation that the proposed method is most useful as a selective correction workflow for visibly problematic polygons, not as a universal replacement for already high-quality building datasets.

\subsection{Remaining shape and post-processing errors}

When excluding the \texttt{SHIFTED} category, the changes in the remaining error composition provide further insight into the types of geometric deficiencies that the workflow was able to correct. During manual labelling, the \texttt{SHAPE\_MISMATCH} category was assigned when a polygon did not clearly represent the visible building shape or represented it only very poorly. The post-processing workflow reduced this category in all study areas, but it remained comparatively frequent in Nepal after processing, where 25.0\% of evaluated buildings still showed a shape mismatch. The reduction of this category is one of the most relevant improvements achieved by the workflow, because these cases included some of the most severe errors in the original dataset.

The \texttt{OVERSIMPLIFIED} category accounted for 13.6\% of the original polygons and decreased to 2.1\% after processing. These polygons generally represented the approximate extent of a building but failed to capture smaller extensions, recesses, roof sections, or other visible geometric details. The original Open Buildings dataset was generated using a semantic segmentation model that classified individual pixels as building or non-building. Building instances were subsequently derived from thresholded prediction masks using connected components and morphological processing \citep{sirko2021continentalscalebuildingdetectionhigh}. Although this approach enables stable building detection at continental scale, the conversion from image information to segmentation masks and finally to vector polygons can limit the representation of small or irregular geometric features.

A visual inspection of the original polygons suggested that many of them had comparatively stable and clean outlines, while smaller details were frequently omitted. However, this should not be interpreted as evidence that the Open Buildings workflow explicitly forced buildings into rectangular or strongly regularized geometries. Rather, the simplified appearance may result from the spatial resolution of the source imagery, the pixel-based segmentation approach, the confidence threshold used to create the masks, and the subsequent extraction of individual building instances. These limitations are particularly relevant for buildings that have developed incrementally over time and therefore contain small extensions, different roof sections, or irregular outlines.

The proposed workflow performed particularly well for this type of error. Instead of predicting all buildings in a larger image tile simultaneously, the visual model could focus on one candidate polygon and its surrounding image context at a time. The geometry was then refined over three consecutive iterations. This allowed the SAM intermediate output to follow the visible building boundary more closely and to recover details that had not been represented in the original polygon. As a result, the SAM output generally provided the highest geometric correspondence with the reference imagery before regularization.

However, the most detailed segmentation result is not automatically the most suitable final GIS geometry. Segmentation masks may contain small boundary fluctuations, local irregularities, or artefacts that are visually plausible but undesirable in a building-footprint dataset. A regularization step was therefore required to improve geometric consistency and usability. Since the preservation of small building features was an important objective of this work, a deliberately non-intrusive regularization approach was selected. The aim was not to force buildings into idealized rectangular shapes, but to remove minor inconsistencies while retaining meaningful extensions, recesses, and other visible details.

The interpretation of the \texttt{MISSING\_PARTS} and \texttt{EXTRA\_PARTS} categories is more difficult. Both categories increased after processing: \texttt{MISSING\_PARTS} from 7.6\% to 15.8\%, and \texttt{EXTRA\_PARTS} from 2.5\% to 10.7\%. This does not necessarily indicate that the workflow generally produced more incomplete or excessive geometries. In the original dataset, many polygons were strongly shifted or showed such poor overall correspondence with the visible building that more specific shape errors could not be assigned reliably. Once the positional alignment and general shape were improved, smaller remaining errors became easier to identify and could be labelled more specifically as \texttt{MISSING\_PARTS} or \texttt{EXTRA\_PARTS}. The increase in these categories may therefore partly reflect a shift from broad error labels toward more precise geometric error descriptions.

The comparison between the SAM intermediate output and the regularized final result shows the central trade-off introduced by post-processing. Across the manually evaluated dataset, the regularization step introduced at least one new geometric error in 16.5\% of the buildings. This was one of the most difficult stages of the workflow to fine-tune. If regularization is too weak, irregularities and segmentation artefacts produced by the visual model remain in the final dataset. If it is too strong, small but meaningful building features are removed, and the resulting polygons become oversimplified. The selected parameterization therefore represents a compromise between geometric stability and fidelity to the visible building outline. While the final regularized polygons were generally more suitable for practical GIS use, the results demonstrate that regularization remains a potential source of new errors and should be evaluated separately from the performance of the visual segmentation model.

Vegetation and roof occlusion were important reasons why errors remained after processing. Buildings with detected trees showed a higher rate of newly introduced post-processing errors than buildings without detected trees. The new-error rate increased from 13.4\% without detected trees to 18.7\% with detected trees, and reached 23.0\% in high tree-context cases (Table~\ref{tab:new_errors_tree_context}). The association was weak, so tree context alone does not explain the failures. Nevertheless, the descriptive pattern is consistent with higher difficulty in vegetation-rich image contexts, especially where roof edges were partly hidden, shadows and vegetation produced ambiguous boundaries, or SAM included surrounding non-building areas in the segmentation.

\section{Role of Foundation Model Components in the Refinement Workflow}

The motivation for using an MLLM for quality assessment was its potential to provide visual reasoning capabilities that are difficult to represent with purely geometric rules. In this workflow, the MLLM was therefore tested for three main tasks: deciding workflow paths, generating semantic error descriptions, and supporting the visual refinement by producing spatial prompt points.

\subsection{MLLM-based quality assessment and workflow decisions}

The MLLM was most useful for coarse workflow decisions. One task was to decide whether there was a corresponding visible building in the image. A second task was to decide whether the full building, or more than 50\% of it, was visible inside the crop, in order to assign the case to the full-house or partial-house workflow. During development, it became clear that the relatively small 8B model worked more reliably when these decisions were split into simpler subquestions instead of being combined in one prompt. This supports the task-decomposition strategy described in the Methods chapter \citep{zhou2022least, Zaid_prompt_decompisiton}.

\begin{itemize}
    \item Building-presence prompt: Is there clear visual evidence of a roof or man-made structure inside the green polygon?
    \item Coverage prompt: Does the green polygon cover more than 50\% of the visible roof, with the complete roof visible inside the image patch?
\end{itemize}

The building-presence prompt showed comparatively reliable behaviour in most study areas. As discussed earlier in Section~\ref{sec:discussion_overall_effectiveness}, the model had particular problems with the recognition of existing buildings in Mozambique. In the other study areas, a short visual comparison suggested more reliable behaviour, although the model still showed difficulties in ambiguous cases, for example with damaged, destroyed, or partly constructed buildings. The second classification prompt, which distinguished full-house and partial-house cases, was less critical. Uncertain cases could be assigned conservatively to the partial-house workflow, where the polygon could still be improved in some cases. In contrast, wrong answers in the building-presence prompt had stronger consequences, because visible buildings could be removed from the pipeline entirely.

A further prompt was tested to let the MLLM compare the original and the post-processed polygon and decide whether the result had improved. This task turned out to be too demanding for the small 8B model. It required the model to compare the input polygon, the image content, and the refined polygon at the same time, and the answers were not consistent enough to be used reliably. This confirms the limitation described in the theoretical background: fine-grained visual comparison is substantially more difficult than coarse visual classification or workflow decisions \citep{wu2024comprehensive}.

The second major MLLM-based task was the generation of semantic error descriptions. As described in Section~\ref{sec:results_mllm_outputs}, two approaches were tested. In both cases, the output was not reliable enough for productive use. When the prompt was formulated very precisely, the model tended to overuse the same predefined error categories. When the prompt was more open, it sometimes produced useful observations, but the descriptions were inconsistent, vague, and often not reproducible. This shows that free-text semantic error description is substantially more difficult than simple workflow decisions. The model had to distinguish between close visual categories such as shifted polygons, missing parts, extra parts, oversimplified outlines, and general shape mismatches. These distinctions require fine-grained spatial reasoning between the visible roof structure and the vector geometry.

This limitation is probably not only a prompt-design problem, but also related to model capacity. The Qwen3-VL technical report presents newer and larger Qwen-VL models as substantially stronger in multimodal reasoning, fine-grained perception, grounding, spatial understanding, and vision-language alignment \citep{bai2025qwen3vltechnicalreport}. Similar model-size effects have also been reported for Qwen-based multimodal models in multi-image and spatial reasoning settings \citep{li2024llavanextinterleavetacklingmultiimagevideo}. These capabilities are directly relevant for semantic geospatial QA, where the model has to relate image content to vector overlays and identify subtle geometric discrepancies. This suggests that a larger or more recent MLLM would likely have improved the consistency and spatial precision of the semantic descriptions. However, model scale alone should not be interpreted as a sufficient solution. Reliable semantic error descriptions would still require structured prompts, predefined error categories, uncertainty handling, and support from deterministic geometric indicators such as IoU, centroid displacement, area difference, or boundary distance.

The third MLLM-based task was the generation of positive and negative prompt points for the visual refinement. Here, the task-decomposition strategy was again useful \citep{zhou2022least,Zaid_prompt_decompisiton}. Instead of asking the model to generate all points in one complex instruction, the point-placement task was split into individual prompts for inside points and outside points. The additional point-in-polygon analysis of manually rated \textit{perfect} full-house cases showed that this component was more reliable than the semantic error description. Across 320 analyzable buildings, 885 of 960 positive points were located inside or on the boundary of the corresponding polygon, resulting in a point-level agreement of 92.19\%. Negative points showed a similar reliability, with 881 of 963 points correctly located outside the polygon, corresponding to 91.48\%. At the stricter building level, however, all positive points were correctly placed for only 80.00\% of buildings. In addition, the third positive point was the weakest point type, with an agreement of 88.44\%. Most errors were small boundary-related deviations, as 51 of the 75 missed positive points were located within 0.5\,m of the polygon boundary.

Overall, the MLLM was more useful for workflow decision-making and approximate spatial assistance than as a reliable semantic evaluator. It helped structure the workflow and provided useful prompt points in many clear full-house cases, but it was not able to generate consistent semantic error descriptions or reliably judge geometric improvement on its own.

\subsection{SAM-based mask generation and vector regularization}

In the developed workflow, SAM was generally not the dominant bottleneck of the geometry refinement. When the input patch and prompt placement were appropriate, the model often produced reliable building masks, even in visually complex settings. This was observed for both point-prompted SAM and text-prompted SAM3 \citep{kirillov2023segment,papersam3}. However, the resulting masks frequently showed irregular, jagged, or aliasing-like boundaries. This is problematic for building extraction, because building footprints are expected to contain straight edges, sharp corners, and structurally meaningful vector geometry.

This limitation is closely related to the raster-based nature of deep learning segmentation masks. SAM predicts binary mask regions, and the final polygon is derived from the pixel-level mask contour. As a result, small local uncertainties along the object boundary can produce visibly irregular outlines after polygonization. The problem was therefore not only whether SAM identified the correct building region, but also how the mask was transformed into a clean vector polygon.

The polygon-based alternatives discussed in the theoretical background illustrate why this distinction matters \citep{girard2021polygonalbuildingsegmentationframe,pix2poly}. In this thesis, SAM remained useful because it provided a flexible and fast segmentation component that could be combined with different prompt types, including points, bounding boxes, input polygons, and text prompts. However, because its output is still mask-based, the subsequent polygon extraction and regularization steps were essential.

A second important limitation was occlusion by vegetation. Since SAM is primarily a visual segmentation model, it segments what is visible in the image. If parts of a roof or building outline are covered by trees, shadows, or other objects, the model cannot reliably infer the hidden footprint \citep{tai2025segment}. In this thesis, no specialized occluded-building adaptation was used. Instead, occlusion was handled through an additional tree-mask segmentation step and post-processing, which attempted to compensate for visually interrupted building outlines.

The role of SAM should therefore be understood in relation to the complete pipeline. SAM was useful for generating local geometry candidates, but its raw output was not directly suitable as final building geometry. The main challenges were prompt selection, occlusion handling, and the conversion of raster masks into clean vector-based building geometries. The mask generation itself was comparatively fast relative to the MLLM-based workflow decisions and point-placement steps. Most of the computational time was caused by point placement, processing decisions, prompt selection, and post-processing rather than by the SAM prediction step itself.

\section{Limitations, Practical Implications, and Future Work}

\textbf{Limitations of the proof-of-concept.}
Before interpreting the results, an important limitation related to the input imagery has to be acknowledged. There was a temporal difference between the individual TIFF images downloaded from OpenAerialMap and the imagery used for the detection of the Google Open Buildings polygons. This was a relevant limiting factor, because the same image source could not be used for both the original building detection and the evaluation carried out in this thesis. This limitation is mainly caused by the restricted availability of very high-resolution imagery across different global regions and by the closed-source nature of the aerial and satellite imagery used for Google Open Buildings. It should also be noted that even in the official Google Open Buildings Explorer, visible shifts between the displayed background imagery and the building polygons can occur, suggesting that different imagery may have been used during the original detection workflow \citep{sirko2021continentalscalebuildingdetectionhigh}.

This aspect is important for the interpretation of the results. The imagery used in this thesis was acquired within a time period of approximately three years around the data used by Google and was not always captured at the same time. Therefore, the evaluated polygons are not necessarily true to the real-world situation at the time of analysis, but rather true to the imagery used in this work. In this thesis, this temporal and positional uncertainty was not treated as a central research question, because this would shift the focus towards remote-sensing image acquisition, image registration, and georeferencing accuracy. Instead, the focus was placed on the quality assessment and selective correction of existing vector data. The following interpretation therefore assumes that the ideal reference would be a building polygon that is geometrically correct with respect to the evaluated imagery, while temporal mismatch, image distortion, and georeferencing uncertainty are treated as external limiting factors \citep{Toutin01052004}. 

Another limitation of the proposed workflow is its computational runtime. A considerable part of the processing time was caused by the MLLM-based requests, especially the two-step semantic error description procedure. As discussed in Section~xx, this component did not provide sufficiently reliable or consistent outputs for productive use, while still adding substantial computational overhead. For an area of approximately xx, the complete workflow required roughly xx minutes to compute. SAM 3 segmentation also contributed to the runtime, especially because each candidate mask was processed over three refinement iterations. However, the segmentation step itself was not the dominant runtime bottleneck compared with the MLLM-based decision and description steps.

A further limitation is the dependency on very high-resolution imagery. The workflow relies on the ability of the MLLM and SAM 3 to visually identify buildings and their boundaries. During development, different image resolutions were tested informally. The workflow appeared to perform reasonably well up to approximately 20\,cm spatial resolution, but performance decreased noticeably in a low-contrast setting with approximately 30\,cm resolution. However, no systematic experiment was conducted to quantify this effect. It is therefore not possible to define a strict minimum resolution threshold based on the results of this thesis. The workflow should consequently be understood as most suitable for areas where high-resolution aerial imagery is available, for example from drone-based local mapping campaigns with reliable georeferencing. Model adaptation, domain-specific fine-tuning, or more specialized training data could potentially reduce this dependency, but this was not investigated in this work.

The attempt to generate reliable semantic error descriptions also has to be considered a limitation. As shown in Section~\ref{sec:results_free_text_descriptions}, the tested MLLM was not able to consistently distinguish between detailed geometric error types such as shifted polygons, missing parts, extra parts, oversimplified outlines, and general shape mismatches. The generated descriptions were either too repetitive when the prompt was constrained, or too vague and inconsistent when the prompt was formulated more openly. Therefore, semantic error description was not suitable as a reliable automated quality-assessment component in the final workflow.


\textbf{Future work.}

