\chapter*{Declaration on the Use of Generative Artificial Intelligence}
\thispagestyle{rmplain}
\label{02_ai_declaration}

I hereby declare that I used generative artificial intelligence tools as supporting aids during the preparation of this Master's thesis. The main tools used were ChatGPT by OpenAI and Codex by OpenAI. These tools were used selectively for language revision, feedback, structuring support, technical assistance, and code-development support during the writing and implementation process.

Generative AI was used for the following purposes:

\begin{itemize}
    \item language support, including grammar correction, academic rephrasing, shortening, and improving the clarity of text passages written by me;
    \item feedback on the structure of chapters, section transitions, and the overall argumentation flow;
    \item support in refining figure captions, table captions, methodological descriptions, and explanatory text for theoretical concepts;
    \item assistance with LaTeX formatting, including tables, figure environments, labels, equations, and citation wording;
    \item conceptual support for checking and improving wording of explanations of technical background topics such as neural networks, convolutional neural networks, transformers, self-attention, foundation models, multimodal large language models, and the Segment Anything Model;
    \item feedback on schematic illustrations, including suggestions for clearer labels, visual structure, and captions. The figures themselves were selected, created, adapted, or redrawn by me;
    \item support in code-related tasks, including debugging suggestions, refactoring ideas, implementation feedback, prompt formulation for code assistants, and explanations of implementation issues;
    \item use of Codex as a code-assistance tool for suggesting edits, identifying potential issues, improving code structure, and supporting the implementation and maintenance of scripts used in the thesis workflow.
\end{itemize}

Generative AI was not used to replace the scientific work, to fabricate data, to generate empirical results independently, or to conduct the manual validation on my behalf. Empirical results were produced by executing the implemented workflow and analysis scripts and were checked, interpreted, and validated by me. The research design, selection of methods, implementation of the workflow, execution of analyses, evaluation of results, interpretation of findings, and final conclusions were carried out by me.

I accept full responsibility for the content, scientific integrity, correctness, and originality of this thesis, including all sections in which generative AI tools were used as support.

\vspace{1.5cm}

